---
title: "Codes"
author: "Alice Zhang"
format: html
editor: visual
---

## Data
### Load Packages

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(janitor)
library(stringr)
library(ggplot2)
library(httr2)
library(knitr)
```

### Retrieve Data

For population data, only 2020-2021 population was available from pset 4 dataset.

```{r}
url <- "https://api.census.gov/data/2021/pep/population"
census_key = "7787d890d72636773b5bc1694aef2bff7ea56237"

request <- request(url) |>
  req_url_query(get = I("POP_2020,POP_2021,NAME"),
                `for` = I("state:*"),
                key = census_key)
response <- request |> req_perform()
pop <- response |> resp_body_json(simplifyVector = TRUE)
```

Get COVID data:

```{r}
api <- "https://data.cdc.gov/resource/pwn4-m3yp.json"

cases_full <- request(api) |>
  req_url_query(`$limit` = 10000000000) |>  
  req_perform() |>
  resp_body_json()|>
  map_df(~ as_tibble(.))
```

## Wrangling

```{r}
#| message: false
#| warning: false

population <- pop |>
  row_to_names(row_number = 1) |>
  as_tibble() |> 
  select(-state) |> 
  rename(state_name = NAME) |> 
  pivot_longer(-state_name, names_to = "year", values_to = "population") |> 
  mutate(year = str_remove(year, "POP_")) |> 
  mutate(across(-state_name, as.numeric)) |> 
  mutate(state = case_when(state_name == "Puerto Rico" ~ "PR",
                           state_name == "District of Columbia" ~ "DC",
                           TRUE ~ state.abb[match(state_name, state.name)]))

cases_cleaned <- cases_full |>
  select(state, end_date, new_cases) |>
  mutate(date = as.Date(end_date, format = "%Y-%m-%d"),
         cases = as.numeric(new_cases)) |>
  select(-end_date, -new_cases)

head(cases_cleaned)
```

# Question 1

Divide the pandemic period, January 2020 to December 2024 into waves. Justify your choice with data visualization.

```{r}
cases_week <- cases_cleaned |>
  mutate(week = floor_date(date, "week")) |> # Aggregate by week
  group_by(state, week) |>
  summarise(total_cases = sum(cases, na.rm = TRUE), 
            .groups = "drop")

# Summarize across states
us_cases <- cases_week |>
  group_by(week) |>
  summarise(total_cases = sum(total_cases, na.rm = TRUE), 
            .groups = "drop")
```

```{r}
wave_periods <- data.frame(
  wave = c("Wave 1", "Wave 2", "Wave 3", "Wave 4", "Wave 5"),
  start = as.Date(c("2020-01-01", "2020-09-01", "2021-06-01",
                    "2022-04-01", "2022-11-01")),
  end = as.Date(c("2020-08-30", "2021-05-31", "2022-03-31", 
                  "2022-10-31", "2023-12-31"))
)

ggplot(us_cases, aes(x = week, y = total_cases)) +
  geom_line(color = "blue", size = 1) +
  geom_rect(data = wave_periods, 
            aes(xmin = start, xmax = end, ymin = 0, ymax = Inf, fill = wave), 
            alpha = 0.2, inherit.aes = FALSE) +
  scale_fill_manual(values = c("Wave 1" = "red", "Wave 2" = "orange", 
                               "Wave 3" = "yellow", "Wave 4" = "green",
                               "Wave 5" = "purple")) +
  labs(title = "COVID-19 Cases Over Time with Waves",
       x = "Week",
       y = "Total Cases",
       fill = "Pandemic Waves") +
  theme_minimal()

```

# Question 2

For each period compute the deaths rates by state. Describe which states did better or worse during the different periods.

```{r, warning=FALSE}
# Assign cases to waves
cases_by_wave <- cases_cleaned |>
  mutate(wave = case_when(
    date >= wave_periods$start[1] & date <= wave_periods$end[1] ~ "Wave 1",
    date >= wave_periods$start[2] & date <= wave_periods$end[2] ~ "Wave 2",
    date >= wave_periods$start[3] & date <= wave_periods$end[3] ~ "Wave 3",
    date >= wave_periods$start[4] & date <= wave_periods$end[4] ~ "Wave 4",
    date >= wave_periods$start[5] & date <= wave_periods$end[5] ~ "Wave 5",
    TRUE ~ NA_character_
  )) |>
  filter(!is.na(wave)) |>
  group_by(state, wave) |>
  summarise(total_cases = sum(cases, na.rm = TRUE), .groups = "drop")

# Include population data 
cases_with_population <- cases_by_wave |>
  left_join(population, by = c("state")) |>
  mutate(case_rate = total_cases / population * 100000)
```

```{r, warning=FALSE}
ggplot(cases_with_population, aes(x = wave, y = case_rate, fill = state)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "COVID-19 Case Rates by State and Wave",
       x = "Wave",
       y = "Case Rate (per 100k)",
       fill = "State") +
  theme_minimal()

# Highlight states with highest and lowest case rates
rates = cases_with_population |>
  group_by(wave) |>
  summarise(
    max_rate_state = state[which.max(case_rate)],
    max_rate = max(case_rate, na.rm = TRUE),
    min_rate_state = state[which.min(case_rate)],
    min_rate = min(case_rate, na.rm = TRUE))

kable(rates)
```

# Question 3

Describe if COVID-19 became less or more virulent across the different periods.

```{r}
# Summarize total cases per wave
cases_by_wave_summary <- cases_with_population |>
  group_by(wave) |>
  summarise(
    total_cases = sum(total_cases, na.rm = TRUE),
    avg_case_rate = mean(case_rate, na.rm = TRUE)
  )
```

# Supplementary

Initial cases vs. time (week) plot for identifying waves.

```{r}
# Plot the entire US cases over time
ggplot(us_cases, aes(x = week, y = total_cases)) +
  geom_line(color = "blue", size = 1) +
  # geom_smooth(method = "loess", span = 0.2, color = "red", se = FALSE) +
  labs(title = "COVID-19 Cases Over Time",
       x = "Week",
       y = "Total Cases",
       subtitle = "Identifying Pandemic Waves (2020–2024)") +
  theme_minimal()

```
